---
title: "HW6RMD"
author: "Matt Schwab"
date: "3/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction:
Team Members: Matt Schwab
GitHub Repository Link: https://github.com/mschwab7/Homework-6.git








Start by loading in the happiness data and importing necessary libraries.
```{r}
happy = readRDS("data/HAPPY.rds")
str(happy)
library(tidyverse)
library(ggplot2)
```

Begin cleaning by replacing certain values with NA
```{r}
temp = happy %in% c("IAP", "DK", "NA")
temp[1: 100]

#Replace values with NA
happy = replace(happy, happy == "IAP", NA)
happy = replace(happy, happy == "DK", NA)
happy = replace(happy, happy == "NA", NA)
View(happy)
```

Now we need to clean the data by converting categorical variables into factors.
Numerical variables must be converted to numerics.We also have to reorder some of the factor variables.

```{r}
#Replace "89 OR OLDER" with 89
happy = replace(happy, happy == "89 OR OLDER", 89)
happy$AGE = as.numeric(happy$AGE)
str(happy)
ggplot(happy, aes(x = AGE)) + geom_bar()
```

Looks like most of the people who were surveyed are in a younger demographic.

```{r}
#Change categorical columns into factors
happy$YEAR = as.factor(happy$YEAR)
```

```{r}
#Change multiple columns at once into factor variables
happy$HAPPY = as.factor(happy$HAPPY)
levels(happy$HAPPY)

happy = happy %>% mutate(SEX = as.factor(SEX), MARITAL = as.factor(MARITAL), 
                 DEGREE = as.factor(DEGREE), FINRELA = as.factor(FINRELA), 
                 HEALTH = as.factor(HEALTH), PARTYID = as.factor(PARTYID),
                 POLVIEWS = as.factor(POLVIEWS))
```

Now we have to reorder some factor variables.

```{r}
levels(happy$YEAR)
```
Year does not have to be reordered

```{r}
levels(happy$MARITAL)
happy$MARITAL = reorder(happy$MARITAL, happy$AGE, fun = mean(), na.rm = TRUE)
levels(happy$MARITAL)
```
Marital should be reordered based on age.

```{r}
levels(happy$DEGREE)
 levels(factor(happy$DEGREE, levels = c("LT HIGH SCHOOL", "HIGH SCHOOL", "JUNIOR COLLEGE","BACHELOR", "GRADUATE")))
levels(happy$DEGREE)
```
Now degree is reordered by prestige.

```{r}
levels(happy$FINRELA)

levels(factor(happy$FINRELA, levels = c("FAR BELOW AVERAGE", "BELOW AVERAGE", "AVERAGE", "ABOVE AVERAGE", "FAR ABOVE AVERAGE")))

levels(happy$FINRELA)
```
Now financial standing is ordered from lowest to highest.

```{r}
levels(happy$HEALTH)
happy$HEALTH = factor(happy$HEALTH, levels = c("POOR", "FAIR", "GOOD", "EXCELLENT"))
levels(happy$HEALTH)

```
Health is now reordered form worst health to best health.

Now everything is reordered. It is time to start exploring the data. But first,
we must save the changed dataset.

```{r}
saveRDS(happy, file = "HW6 rds")
```

```{r}
happy %>% ggplot(aes(x = HAPPY)) + geom_bar() + facet_wrap(~DEGREE)
happy %>% ggplot(aes(x = HAPPY)) + geom_bar() + facet_wrap(~MARITAL)
happy %>% ggplot(aes(x = HAPPY)) + geom_bar() + facet_wrap(~SEX)
happy %>% ggplot(aes(x = HAPPY)) + geom_bar() + facet_wrap(~FINRELA)
```
We made a couple charts to explore the data. Now it is time to dig deeper.We want to find out how happiness is related to multiple variables. To do this, we want to treat happiness as a numeric variable, where a higher value corresponds to a higher level of happiness.

```{r}
levels(happy$HAPPY)
```

```{r}
happy %>% group_by(FINRELA, HEALTH) %>% summarise(m.happy = mean(as.numeric(HAPPY), na.rm = TRUE))
happy.fin = happy %>% group_by(FINRELA, HEALTH) %>% summarise(m.happy = mean(as.numeric(HAPPY), na.rm = TRUE))
happy.fin %>% ggplot(aes(x = FINRELA, fill = HEALTH)) + geom_bar(aes(weight = m.happy),
                                                              position = position_dodge())
```
Strangely enough, it seems that people with worse health tend to be more happy, and people with better health seem to be less happy. It looks like happiness levels don't change across health when financial relationship increases. So people in poor health with lots of money are about as happy as people in poor health with little money. 


```{r}
happy %>% group_by(MARITAL, FINRELA) %>% summarise(m.happy2 = mean(as.numeric(HAPPY), na.rm = TRUE))
happy.mar = happy %>% group_by(MARITAL, FINRELA) %>% summarise(m.happy2 = mean(as.numeric(HAPPY), na.rm = TRUE))
happy.mar %>% ggplot(aes(x = FINRELA, fill = MARITAL)) + geom_bar(aes(weight = m.happy2),
                                                                  position = position_dodge())

```

It looks like people who are married tend to be the happiest, while people who are separated tend to be the least happy.It seems like for all people not including married people, happiness levels actually decline as income increases. This may suggest that as people get more money, they long for more companionship, and if they don't have it they are not happy. 


```{r}
happy %>% summarise(m.happy = mean(as.numeric(HAPPY), na.rm = TRUE))
happy %>% group_by(SEX) %>% summarise(m.happy = mean(as.numeric(HAPPY), na.rm = TRUE))
```
In general females seem to be slightly more happy than males, without taking 
into consideration any other factors.

```{r}
happy %>% summarise(m.happy = mean(as.numeric(HAPPY), na.rm = TRUE))
happy %>% group_by(FINRELA) %>% summarise(m.happy = mean(as.numeric(HAPPY), na.rm = TRUE))
```

Does money on its own "buy" happiness? The data remains a bit unclear. The highest mean happiness is for people with far below average financial standing.
This is then followed by people with far above average finances. This goes to show that money alone does not relate to happiness; happiness is dependent on the combination of the factors, not one factor on its own.


